---
title: "SequenceAnalysisEB"
author: "Cardenas-Rivera, A."
date: "2026-05-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Bioinformatics Advances

## Practical Assignment: Sequence Analysis of Epidermolysis Bullosa (EB) Genes Using R

### Learning Objectives

By the end of this practical, students will be able to:

1. Import and manipulate DNA and protein sequences in R.
2. Compute descriptive and biophysical statistics from protein sequences.
3. Identify regulatory motifs and splice-site signals.
4. Characterize hydrophobic regions and predict transmembrane domains.
5. Perform multiple sequence alignment (MSA).
6. Construct and interpret phylogenetic trees.
7. Integrate biological interpretation with computational results.

---

# Background

Epidermolysis Bullosa (EB) is a group of inherited skin disorders characterized by skin fragility and blister formation. Several genes are frequently implicated, including:

* KRT5 (Keratin 5)
* KRT14 (Keratin 14)
* COL7A1 (Collagen VII Alpha 1)

In this practical, you will analyze DNA and protein sequences from EB-associated genes using R.

---

# Required Packages
```{r}
# Install packages

#install.packages(c("seqinr", "ape"))
#install.packages("rentrez")

#if (!requireNamespace("BiocManager"))
#  install.packages("BiocManager")

#BiocManager::install(c("Biostrings", "msa"))


#Load libraries:

library(seqinr)
library(Biostrings)
library(msa)
library(ape)
library(rentrez)
```

---

# Phase 1: Data Acquisition and Descriptive Statistics

## Objective

Generate a biophysical profile of an EB protein sequence.

### Biological Question

What are the basic biochemical properties of the protein?

### Tasks

1. Download the protein FASTA sequence of KRT14 from NCBI.
2. Import the sequence into R.
3. Calculate amino acid frequencies.
4. Calculate molecular composition.
5. Determine the isoelectric point.

### Code

#### Load sequence

```{r}
# load from local file
# krt14 <- read.fasta("KRT14.fasta")

# load from NCBI repository
krt14 <- entrez_fetch(db = "nuccore", 
                            id = "NM_000526", 
                            rettype = "fasta")
krt14
```

#### Convert to read.fasta format

```{r}
krt14 <- seqinr::read.fasta(file = textConnection(krt14), 
                                      seqtype = "DNA", 
                                      as.string = TRUE)
summary(krt14)
```

#### Get sequence

```{r}
# as a separate character vector (jby default)
seqKrt14 <- getSequence(krt14[[1]])     # extract only sequence
seqKrt14                  # print sequence
table(seqKrt14)           # contingency table of absolute values
prop.table(table(seqKrt14))# contingency table of relative values

# as a string
seqKrt14 <- getSequence(krt14[[1]], as.string = T)   # extract sequence as STRING
seqKrt14 <- unlist(seqKrt14)           # unlist the sequence
seqKrt14                               # print the sequence

```

### Import a protein sequence

```{r}
# load from NCBI repository
kRT14 <- entrez_fetch(db = "protein",                      # Watch out!!! change DB 
                            id = "NP_000517", 
                            rettype = "fasta")

# convert to read.fasta format
kRT14 <- seqinr::read.fasta(file = textConnection(kRT14), 
                                      seqtype = "AA",       # Watch out!!! change type
                                      as.string = TRUE)
# extract sequence
seqKRT14 <- unlist(getSequence(kRT14[[1]])) 

table(seqKRT14)

AAstat(seqKRT14)
```

### Questions

1. Which amino acids are most abundant?
2. What is the predicted isoelectric point?
3. Does the amino acid composition suggest a structural protein?

---

# Inline Glossary

### FASTA

Standard text format used to store biological sequences.

### Vignette

A tutorial document included with many R packages that demonstrates package functionality using real examples.

---

# Phase 2: Pattern Discovery (Regulatory and Splicing Sites)

## Objective

Identify biologically relevant sequence motifs.

### Biological Question

Can we locate potential promoter and splice-site elements?

---

## Part A: Regulatory Motifs

Search for TATA boxes.

### Code

#### Download sequence and extract sequence

```{r}
# download sequnce from NCBI repository
gnmKrt14 <- entrez_fetch(db = "nuccore", 
                         id = "NG_008624.2", 
                         rettype =  "fasta")

# TIP: If you have a connection issue (lag connection)
# Increase the connection timeout to 60 seconds (60000 milliseconds)
options(timeout = 60)

# Explicitly write the text to the exact file path expected
write(gnmKrt14, file = file.path(dirname(getwd()),"data","KRT14_genomic.fasta"))

# Import from local location 
dna <- readDNAStringSet(file.path(dirname(getwd()),"data","KRT14_genomic.fasta"))

# Convert to sequence
sequence <- dna[[1]]    # unlist 
sequence                # print the sequence

```

#### Find TATA box

A common promoter pattern located about *25* to *30* base pairs upstream of the start site. The consensus sequence is universally: *TATAAA*

```{r}
matchPattern("TATAAA", sequence)
```

### Alternative Motif

#### Search for GC-rich promoter motifs.

```{r}
matchPattern("GGGCGG", sequence)
```

#### Search for a CAAT box
Often found around *-70* to *-80* base pairs upstream, usually following the pattern: *GGCCAATCT*

```{r}
matchPattern("GGCCAATCT", sequence)
```

Change parameters to make it a litle flexible

```{r}
matchPattern("GGCCAATCT", sequence, max.mismatch = 1)
```

### Questions

1. How many TATA motifs were found?
2. Are they located upstream of the coding sequence?

---

## Part B: Splicing Sites

Search for canonical GT/AG splice junctions.

### Code

```{r}
library(Biostrings)

matches_GT <- matchPattern("GT", sequence) # use vmatchPattern() for multiple sequences

matches_AG <- matchPattern("AG", sequence)
```

Using regular expressions:

```{r}
gregexpr("GT.*?AG",
         as.character(sequence))
```

### Questions

1. How many potential splice sites were identified?
2. Why might some predictions be false positives?

---

# Inline Glossary

### Motif

A short recurring DNA sequence associated with a biological function.

### Regular Expression

A pattern-matching language used to identify specific text or sequence structures.

---

# Phase 3: Biophysical Profiling

## Objective

Identify hydrophobic regions and potential transmembrane domains.

### Biological Question

Does the protein contain membrane-spanning regions?

---

## Hydropathy Analysis

Use the Kyte-Doolittle hydropathy scale.

### Code

```{r}
# 1. Install and load required libraries
#if (!requireNamespace("Biostrings", quietly = TRUE)) {
#  if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
#  BiocManager::install("Biostrings")
#}
library(Biostrings)

# 2. Define the Kyte-Doolittle hydrophobicity scale vectors:
kd_scale <- c(A=1.8, R=-4.5, N=-3.5, D=-3.5, C=2.5, Q=-3.5, E=-3.5, G=-0.4, H=-3.2, 
              I=4.5, L=3.8, K=-3.9, M=1.9, F=2.8, P=-1.6, S=-0.8, T=-0.7, W=-0.9, 
              Y=-1.3, V=4.2)

# 3. Define hydrophobicity
hydro <- kd_scale[unlist(strsplit(seqKRT14,""))]

plot(hydro, type="l")
```

Smooth the plot

```{r}
#install.packages("zoo")
library(zoo)

hydro.fil <- rollapply(hydro, width = 20, FUN = mean, fill = NA, align = "center")
plot(hydro.fil, type="l")
```
To identify a potential transmembrane (TM) $\alpha$-helical motif using the Kyte-Doolittle scale, a sliding window average of above *1.6* to *1.8* is generally required. This hydrophobic stretch must typically be sustained for 19 to 21 consecutive amino acids, which is the required length to span the hydrophobic core of a cell membrane
---


### Questions

1. Are there strongly hydrophobic regions?
2. Would you predict transmembrane domains?

---

# Inline Glossary


### Hydropathy

A measure of amino acid hydrophobicity or hydrophilicity.

---

# Phase 4: Multiple Sequence Alignment

## Objective

Identify conserved regions across species.

### Biological Question

Which regions of the protein are evolutionarily conserved?

---

## Tasks

Download orthologous KRT14 protein sequences from:

* Human
* Mouse
* Rat
* Zebrafish
* Cow

---

## Code

#### Import data set

```{r}
seqs <- readAAStringSet(filepath = file.path(dirname(getwd()),"data","krt14_homolo.fasta"))

```

#### Alignment

```{r}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("msa")

library(msa)

alignment <- msa(
  seqs,
  method = "Muscle"
)

alignment
```

#### Visualize:

```{r}
msaPrettyPrint(
  alignment,
  output = "pdf",
  file = file.path(dirname(getwd()),"plot","alignment.pdf"),
  showNames = "left",
  showLogo = "top",
  shadingMode = "similar"
)


# if you get a error with latex pdf
#install.packages("tinytex") # DO NOT restart session for now
#tinytex::install_tinytex()

```

#### Finding the consensus sequence

```{r}
cm <- consensusMatrix(alignment)         # define the consensus matrix
cns <- consensusString(alignment)
cns
```

#### Plot a specific region of interest in the sequence

```{r}
# install.packages("ggseqlogo", dependencies = T)


library(Biostrings)
library(ggseqlogo)


# Convert to character vector
seqs <- as.character(alignment)

# Plot sequence logo
ggseqlogo(seqs, method = "prob")        # test outside the markdown

```

method = "bits"
stack height depends on conservation
conserved positions become taller
noisy positions shrink

This is the classic publication-style sequence logo.


```{r}
ggseqlogo(seqs, method = "bits")

```

using ggplot to improve output


```{r}
library(ggplot2)

ggseqlogo(seqs, method = "bits") +
  theme_bw(base_size = 14) +
  ggtitle("Protein sequence logo")

```

Make a crop of the consensus sequence

```{r}
library(ggseqlogo)
# convert to AAStringSet
alignment_set <- unmasked(alignment)

# Crop alignment positions
cropped_alignment <- Biostrings::subseq(alignment_set, start = 350, end = 360)
# Convert to character vector
seqs_crop <- as.character(cropped_alignment)

# Plot sequence logo
ggseqlogo(seqs_crop, method = "prob") +
  theme_bw(base_size = 14) +
  ggtitle("Protein sequence logo")
```

```{r}
# Plot sequence logo
ggseqlogo(seqs_crop, method = "bit") +
  theme_bw(base_size = 14) +
  ggtitle("Protein sequence logo")
```


### Questions

1. Which regions show the highest conservation?
2. Are disease-associated residues located in conserved regions?

---

# Inline Glossary

### Ortholog

Genes in different species derived from a common ancestral gene.

### Multiple Sequence Alignment (MSA)

Alignment of several sequences to identify conserved regions.

### Wrapper

A function that simplifies the use of a more complex software package.

---

# Phase 5: Evolutionary Inference

## Objective

Reconstruct evolutionary relationships among orthologs.

### Biological Question

How have EB genes evolved across species?

---

## Distance Matrix

```{r}
alignment_matrix <- as.matrix(alignment)

distances <- dist.alignment(
  as.alignment(alignment_matrix),
  "identity"
)

distances
```

---

## Neighbor-Joining Tree

```{r}
tree <- nj(distances)

plot(tree,
     main="KRT14 Evolutionary Tree")
```

Bootstrap support:
The boot.phylo() function from the ape package performs bootstrap support analysis for phylogenetic trees.
It estimates how strongly the data support each branch (clade) in your phylogenetic tree.

```{r}
boot.phylo(tree,
           alignment_matrix,
           function(xx)
             njs(dist.aa(xx)))
```

### Questions

1. Which species is closest to humans?
2. Does the tree reflect known mammalian evolution?
3. Are there any unexpected relationships?

---

# Final Report

Prepare a short report (3–5 pages) including:

1. Introduction
2. Methods
3. Results
4. Biological interpretation
5. Figures and plots
6. Conclusions

---

# Bonus Challenge

Select one pathogenic mutation from ClinVar in KRT14 or COL7A1 and discuss:

* Whether it occurs in a conserved region.
* Whether it alters protein hydrophobicity.
* Whether it may affect splicing.
* Its potential impact on protein function.
